Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-02 text/discussion.qmd
first run through of the discussion section finished

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide list-based marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

As an owner-operated EM consultancy, Royall was the market leader in purchasing student lists on behalf of clients to help realize enrollment goals. Under Vista, EAB tranformed the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours, which can be conceptualized as an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading provider of free college search engine, which is an engine for leads. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista Eqity Partners -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. At the same time, EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform. EAB combined all of these acquisitions into Enroll360:

> We spent the last couple of years creating a connected recruitment ecosystem that allows enrollment leaders to keep pace with students as they pursue these increasingly digital journeys to college. This work led us to join forces with several leading companies: Cappex, Intersect, Wisr, and YouVisit\ldots Individually, each solution can solve important challenges at various stages of the enrollment funnel\ldots By bringing these capabilities together, our vision is to reinvent how enrollment leaders reach their goals [@koppenheffer_2021]

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50). Thus, the goal of lists was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board *Student Seardh Service* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, Investigative reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. A document from the University of Utah to procure an Intersect subscription explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:

@@

We argue that an under-studied and under-regulated third-party EM industry has negative consequences for colleges and for students. Accross multiple student list products, we see a growing number of filters that recommend which prospects to target based on the beheavior of previous cohorts of students. The ACT/Encoura "Enrollment Predictor" filter enables colleges to filter prospects based on the predicted probability of enrollment, which is created based on analyzing the enrollment decisions of previous cohorts of students [schmidt_2019]. Based on the Claritas/PRIZM market segmentation system that categorized zip codes into groups useful for merchants [@RN4957], The College Board Geodemographic Segment filters categorizes each Census Tract into one of about two dozen clusters based on the college-going behavior of previous cohorts [@RN4565]. A @vista_eab_ml video promoting EAB's "Pipeline Analytics" product to investors states, "what we are doing is using artificial intelligence and machine learning to help schools build list of students may be more more likely to enroll and persist in their university." The product "identifies students that would be a good match for a school based on their similarity to students who have succeeded there before and then provides them [the students] information about those institutions so that they choose to apply and ultimately enroll." Each of these products are based on the underlying logic of correlations and homophily discussed by @RN4975 and applied by @RN4982. These products identify correlations in a historic snapshot of students and apply those relationships to make recommendations about which prospects to target in the present. As such, all of these technologies amplify the effect of previous inequality in student demand, which is itself a consequence of structural inequality in educational opportunity.

Market dynamics in the EM industry may also affect competition, prices, and choice. To the extent that colleges need to buy expensive software to recruit prospects contained in an associated database, those costs will ultimately be passed on to students in the form of higher prices are reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that don't have vertical inputs -- e.g., control over proprietary pools of prospects -- cannot compete. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that are willing and able to pay for subscriptions, which raises policy concerns about student choice.

@@

Regulation of the higher education industry should be informed by research that investigates the "totality of relevant actors" [@RN527, p. 148] in the organizational field. The paucity of research about third-party vendos that structure college access is substantially due to difficulty obtaining data for empirical analyses.

@@

We argue that sociologists and adjacent scholars should expand the domain of the education research knowledge infrastructure to include third-party vendors in enrollment management, and edtech more broadly. This is a difficult task. Student unit record data owned by district, state, and federal agencies is simultaneously granular -- can evaluate a particular policy -- and macro -- can speak to broad trends in the education sector. By contrast, obtaining granular data about third-party products in education is difficult because these data are owned by private interests that do not want researchers to interrogate their products.

We see two broad avenues for research: case studies involving primary data collection -- including this study -- which tend to have a granular focus on particular products or vendors; and secondary data analysis, which may have a more macro focus. Case study research traditionally relies on interviews, ethnographies, and document analysis. Another approach to data collection is to issue records requests to public entities that contract with third-party entities. For example, @RN5039 analyze public university contracts with online program managers (OPMs). We issued public records requests to obtain tabular data about student list purchases. However, this process was labor intensive, requiring the pro bono efforts of several law firms. Public records requests seem better suited to obtaining contracts. Text analysis of publicly available websites and social media provides another avenue for data collection. Here, data collection is efficient but data processing is often labor intensive. 
Data collected for case study research typically does not meet the knowledge infrastructure standard of repeated, ongoing data collection that can be the foundation for scalable research. Nevertheless, case study research is important for granular analyses of particular products -- like Intersect, Enroll360,  that yield important insights about underlying mechanisms about how vendors structure college access.

A growing set of "market intelligence" data providers serve the information needs of the investment community by providing data on private firms and deals involving private firms. These databases may provide the foundation for scalable research on the role of private firms and private equity investors in education. These databases tend to include three "types" of data: (1) *firm-level data* (e.g., location, employees, financials); (2) *acquisitions*, encompassing who acquired whom and resulting ownership structure); and (3) *investments*, which include amounts/valuations/stake percentages of private equity investments. Some providers provide all three types of data (e.g., [*S&P Capital IQ*](https://www.spglobal.com/market-intelligence/en), [*LSEG Workspace*](https://www.lseg.com/en)). Other providers specialize. For example, [*PitchBook*](https://pitchbook.com/) is a leading provider of data about investments in private companies by private equity firms, while [*Orbis*](https://login.bvdinfo.com/R1/Orbis) is the leading provider of firm-level data for non-U.S. private firms [@RN5038]. Interestingly, the database subscriptions operate similar to contemporary student list products; an interactive, user-facing platform is wrapped around several sources of proprietary data. University libraries provides full or partial access to some subscriptions, while others must be purchased. These databases are often utilized for articles in management journals that focus on private equity markets [CITE], or patterns of acquisitions [CITE]. Considering the growing share of economic activity that  [HOW THESE DATABASES HAVE BEEN USED; HOW THEY MAY BE USED]

>>>>

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

As an owner-operated consultancy, Royall was the market leader in purchasing student lists from student list vendors on behalf of clients. Under Vista, EAB entered the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours. YouVisit is an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading free college search engine, which is an engine for leads in that students provide contact details and other information. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform. EAB combined these acquisitions into create [Enroll360](https://eab.com/solutions/enroll360/):

> We spent the last couple of years creating a connected recruitment ecosystem that allows enrollment leaders to keep pace with students pursuing increasingly digital journeys to college. This work led us to join forces with several leading companies: Cappex, Intersect, Wisr, and YouVisit\ldots Individually, each solution can solve important challenges at various stages of the enrollment funnel\ldots By bringing these capabilities together, our vision is to reinvent how enrollment leaders reach their goals [@koppenheffer_2021]

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board *Student Search Service* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. A procurement document from the University of Utah explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:

@@

An under-studied and under-regulated third-party EM industry has negative consequences for colleges and for students. Accross multiple student list products, we see a growing number of filters that recommend which prospects to target based on the beheavior of previous cohorts of students. The ACT/Encoura "Enrollment Predictor" filter enables colleges to filter prospects based on the predicted probability of enrollment, which is created based on analyzing the enrollment decisions of previous cohorts of students [@schmidt_2019]. College Board Geodemographic Segment filters categorizes each Census Tract into one of about two dozen clusters based on the college-going behavior of previous cohorts [@RN4565] and is based on the Claritas/PRIZM market segmentation system that categorized zip codes into groups useful for merchants [@RN4957]. An investor-facing @vista_eab_ml video promoting EAB's "Pipeline Analytics" product states, "we are using artificial intelligence and machine learning to help schools build list of students that may be more more likely to enroll and persist in their university." The product "identifies students that would be a good match for a school based on their similarity to students who have succeeded there before and then provides them [the students] information about those institutions so that they choose to apply and ultimately enroll." Each of these products are based on the underlying logic of correlations and homophily discussed by @RN4975 and applied by @RN4982. These products identify correlations in a historic snapshot of students and apply those relationships to make recommendations about which prospects to target in the present. As such, all of these technologies amplify the effect of previous inequality in student demand, which is itself a consequence of structural inequality in educational opportunity.

Market dynamics in the EM industry raise policy concerns about tuition prices, student choice, and competition in the market for EM consultancies
may also affect competition, prices, and student choice. To the extent that colleges must buy expensive software to recruit prospects contained in an associated database, this expenditure will ultimately be passed on to students in the form of higher prices or reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that do not control veritial inputs may be unable to compete with companies that have acquired proprietary pools of prospects. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that are willing and able to pay for subscriptions, which raises policy concerns about student choice.

@@

Regulation of the higher education industry should be informed by research that investigates the "totality of relevant actors" [@RN527, p. 148] in the organizational field. Unfortunatel, scholarship salient to education policy and regulation have ignored two, related economic transformations. First, third-party vendors offer software platforms that perform core tasks of organizations, including schools and colleges [@list_biz]. Second, private firms comprise a growing share of economic activity [@RN5037; @RN5038]. Most firms in EM and the broader edtech sector are private. However, private firms face fewer disclosure requirements than publicly traded firms. The paucity of research about third-party vendors that structure college access is substantially due to difficulty obtaining data for empirical analyses. The research community must figure out how to systematically investigate the role of private firms and investors in education, so that we can hold private interests accountable for education policy goals.

@@

We argue that sociologists and adjacent scholars should expand the domain of the education research knowledge infrastructure to include third-party vendors in enrollment management, and edtech more broadly. This is a difficult task. Student unit record data owned by district, state, and federal agencies is simultaneously granular -- can evaluate a particular policy -- and macro -- can speak to broad trends in the education sector. By contrast, obtaining granular data about third-party products in education is difficult because these data are owned by private interests that do not want researchers to interrogate their products [@RN4774; @RN4842]. We see two broad avenues for research: case studies involving primary data collection -- including this study -- which tend to have a granular focus on particular products or vendors; and secondary data analysis, which may have a more macro focus.

Although case study research often relies on interview and ethnographic data, another approach to data collection is to issue records requests to public entities that contract with third-party entities. For example, @RN5039 analyze public university contracts with online program managers (OPMs). The present manuscript is an example of case study research. We issued public records requests to obtain tabular data about student list purchases. However, this process was labor intensive, requiring the pro bono efforts of several law firms. Public records requests seem better suited to obtaining contracts. Qualitative or computational text analysis of publicly available websites and social media provides another avenue for data collection. Here, data collection is efficient but data processing is often labor intensive. Like investigative reporting [@markup_naviance], case study research can produce granular analyses of particular products (e.g., Intersect, Enroll360) that yield insight into the underlying mechanisms of how third-party vendors structure college access. However, data collected for case study research typically does not meet the knowledge infrastructure standard of repeated, ongoing data collection that can be the foundation for scalable research.

We see subscription databases about private markets as having the potential to create a knowledge infrastructure for scalable research on the role of private firms and private equity funders in education, and also their role in other domains of interest to sociology. A growing set of "market intelligence" data providers serve the information needs of the investment community by providing data on private firms and deals involving private firms. These databases tend to include three broad "types" of data: (1) *firm-level data* (e.g., location, employees, financials); (2) *acquisitions*, encompassing who acquired whom and resulting ownership structure); and (3) *investments*, which include amounts/valuations/stake percentages of private equity investments. Some providers provide all three types of data (e.g., [*S&P Capital IQ*](https://www.spglobal.com/market-intelligence/en), [*LSEG Workspace*](https://www.lseg.com/en)). Other providers specialize. For example, [*PitchBook*](https://pitchbook.com/) is a leading provider of data about investments in private companies by private equity firms, while [*Orbis*](https://login.bvdinfo.com/R1/Orbis) is a leading provider of firm-level data for non-U.S. private firms. These databases are often utilized in journals of finance or management that consider private equity markets or patterns of acquisitions [e.g., @RN5044; @RN5038]. Interestingly, the database subscriptions operate similar to contemporary student list products; an interactive, user-facing platform is wrapped around several sources of proprietary data. University libraries provides full or partial access to some subscriptions, while others must be purchased.

Although scholars cannot share subscription data, they can share code to process and analyze these data. We recommend that scholars develop panel datasets examining change over time in the firms operating in particular industries and acquisitions and private equity investments in those industries. For example, empirical scholarship can examine change over time in concentration in the enrolment management industry and its transformation from an owner-operated to a largely private-equity owned industry. Subsequent analyses could investigate the portfolio of large owners, their ties to other industries -- such as the direct-provider for-profit college market -- and the vertical inputs/technologies they are investing in.
